Implementing simplified INLA

### 2022-09-29 src/dev_sinla/sinla.Rmd
Iterate on SINLA development. Still trying to get TMB Laplace approximation to match with (more) manually calculated one

---
title: "Implementing simplified INLA"
author:
- name: Adam Howes
output:
  html_document:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
    df_print: paged
    code_folding: show
    theme: lumen
abstract: |
  **Background** Wood (2020) simplify the INLA method by approximating latent field posterior marginals using a reduced cost Laplace approximation which does not rely on sparsity assumptions.
  **Task** We implement the method in R making use of Template Model Builder for Gaussian approximations.
---

@@

The AGHQ/ELGM approximation to the joint posterior is
$$
\tilde p(x \, | \, y) = \sum_z \tilde p_\text{G}(x \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z).
$$
This can be evaluated at any point, though in practice we work with it by drawing samples $x \sim p_\text{G}(x \, | \, \theta, y)$ using a method from sampling from a Gaussian given its mean and precision from Rue.
For the $i$th marginal, we just take the appropriate samples from the joint.
The approximation is therefore
$$
\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{G}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z),
$$
where $p_\text{G}(x_i \, | \, \theta, y)$ just takes the $i$th elements of the mean vector and $(i, i)$th element of the covariance matrix.

@@

We want to create a new approximation swapping out $\tilde p_\text{G}(x_i \, | \, \theta, y)$ from the above for the simplified INLA approximation
$$
\tilde p(x_i \, | \, \theta, y) = \sum_z \tilde p_{\text{SINLA}}(x_i \, | \, \theta, y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z).
$$

@@

We create an objective function `f` (see Appendix for `TMB` template) to evaluate $p(x, \theta, y)$ (without any Laplace approximations!):

```{r message=FALSE, warning=FALSE}
param <- list(
  beta_prev = 0,
  phi_prev = rep(0, data$n),
  log_sigma_phi_prev = 0
)

@@

@@

```{r message=FALSE, warning=FALSE, results=FALSE}
#' To store obj
template <- list()

@@

  template[[i]] <- obj
}
```

@@

```{r message=FALSE, warning=FALSE}
i <- 1
obj <- template[[i]]

mean <- with(obj$env, last.par[random])
length(mean) #' The length is 30 - 1
mean

Q <- obj$env$spHess(with(obj$env, last.par), random = TRUE)
dim(Q) #' With dimension [(30 - 1) x (30 - 1)]
image(Q)
```

@@

We can reproduce the output of `obj$fn` using the mean and precision matrix, together with the function `f`.
Recall that `f` takes as input the complete $(x, \theta)$ ($y$ is already fixed).
For example:

```{r}
f$fn(unlist(param))
```

We update `param` to include $\hat{\mu}_{-i}(x_i, \theta)$:

```{r}
param$phi_prev[-i] <- mean
param

```

Such that the numerator of the Laplace approximation is (n.b. I think this is on the log-scale, which is where we will do the calculations):

```{r}
numer <- f$fn(unlist(param))
numer
```

And the denominator of the Laplace approximation (on the log-scale) is:

```{r}
denom <- log(sqrt(det(2 * pi * Q)))
denom
```

```{r}
numer - denom
```

>>>>

---
title: "Implementing simplified INLA"
author:
- name: Adam Howes
output:
  html_document:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
    df_print: paged
    code_folding: show
    theme: lumen
abstract: |
  **Background** Wood (2020) simplify the INLA method by approximating latent field posterior marginals using a reduced cost Laplace approximation which does not rely on sparsity assumptions.

  **Task** We implement the method in R making use of Template Model Builder for Gaussian approximations.
---

@@

# Further details on what we would like to achieve

This section is where I am working things out about what we're aiming for.
It likely won't follow much narrative structure for now.

The AGHQ/ELGM approximation to the joint posterior is
$$
\tilde p(x \, | \, y) = \sum_z \tilde p_\text{G}(x \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z),
$$
where $\theta(z) \in \Theta$ and $w_k(z) \in \mathbb{R}$ are quadrature nodes and weights.
In particular, let
$$
\mathcal{Q}(1, k) = \{z_d \in \mathbb{R}: H_k(z_d) = 0\}
$$
where $H_k(z_d)$ are the Hermite polynomials for $k \in \mathbb{N}$, and
$$
w_k(z_d) = \frac{k!}{H_{k + 1}(z_d)^2 \times \phi(z_d)}, 
$$
Define $\mathcal{Q}(d, k) = \mathcal{Q}(1, k)^d$ with nodes $z$, of which there will be $k^d$.
Then $\theta(z) = \hat \theta + Lz$ where $\hat \theta$ is the mode and $L$ is the lower Cholesky triangle of the curvature such that $LL^\top = H^{-1}$.

This can be evaluated at any point, though in practice we work with it by drawing samples $x \sim p_\text{G}(x \, | \, \theta, y)$ using a method from sampling from a Gaussian given its mean and precision from Rue.
For the $i$th marginal, we just take the appropriate samples from the joint.
The approximation is therefore
$$
\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{G}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z),
$$
where $p_\text{G}(x_i \, | \, \theta, y)$ just takes the $i$th elements of the mean vector and $(i, i)$th element of the covariance matrix.

@@

We want to create a new approximation swapping out $\tilde p_\text{G}(x_i \, | \, \theta, y)$ from the above for the simplified INLA approximation $p_{\text{SINLA}}(x_i \, | \, \theta, y)$ as follows
$$
\tilde p(x_i \, | \, \theta, y) = \sum_z \tilde p_{\text{SINLA}}(x_i \, | \, \theta, y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z).
$$

@@

We create an objective function `f` (see Appendix for `TMB` template) to evaluate $- \log p(x, \theta, y)$ (without any Laplace approximations!):

```{r message=FALSE, warning=FALSE}
param <- list(
  beta_prev = -2,
  phi_prev = rep(0, data$n),
  log_sigma_phi_prev = -1
)

@@

Optimise `h`:

```{r}
its <- 1000

opt <- nlminb(
  start = h$par,
  objective = h$fn,
  gradient = h$gr,
  control = list(iter.max = its, trace = 0)
)
```

Get the parameters of the Gaussian approximation:

```{r}
mean <- with(h$env, last.par[random])
length(mean) #' The length is 30
mean

Q <- h$env$spHess(with(h$env, last.par), random = TRUE)
dim(Q) #' With dimension [30 x 30]
image(Q)

input_param <- param
h$fn(input_param[c("beta_prev", "log_sigma_phi_prev")])
```

Try to reproduce by input to `f`:

```{r}
input_param$phi_prev <- mean
numer <- -1 * f$fn(unlist(input_param))
numer

denom <- 0.5 * log(det(2 * pi * Q))
denom
```

On the negative log-likelihood scale, the number we get is:

```{r}
-1 * (numer - denom)
```

@@

```{r message=FALSE, warning=FALSE, results=FALSE}
objs <- list()
means <- list()
Qs <- list()

@@

  #' Parameters of the Gaussian approximation
  mean <- with(obj$env, last.par[random])
  Q <- obj$env$spHess(with(obj$env, last.par), random = TRUE)

  objs[[i]] <- obj
  means[[i]] <- mean
  Qs[[i]] <- Q
}
```

@@

```{r message=FALSE, warning=FALSE}
i <- 1
obj <- objs[[i]]
mean <- means[[i]] #' The length is 30 - 1
mean

Q <- Qs[[i]]
dim(Q) #' With dimension [(30 - 1) x (30 - 1)]
image(Q)
```

@@

We should be able to reproduce the output of `obj$fn` using the mean and precision matrix, together with the function `f`.
Recall that `f` takes as input the complete $(x, \theta)$ ($y$ is already fixed).
For example:

```{r}
param
f$fn(unlist(param))
```

We update `param` to include $\hat{\mu}_{-i}(x_i, \theta)$, so that it's only element `r i` of `phi_prev` which remains as before:

```{r}
param$phi_prev[-i] <- mean
param
```

The numerator of the Laplace approximation is (n.b. this is on the log-scale, which is where we will do the calculations):

```{r}
numer <- -1 * f$fn(unlist(param))
numer
```

And the denominator of the Laplace approximation (again on the log-scale) is:

```{r}
denom <- 0.5 * log(det(2 * pi * Q))
denom
```

```{r}
-1 * (numer - denom)
```
